---
title: "A worked example with the ChrAccR example data"
author: "Irem B. Gunduz"
package: chromTFR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{A worked example with the ChrAccR example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
haveData <- all(vapply(
    c("ChrAccR", "ChrAccRex", "methylTFRAnnotationHg38"),
    requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(eval = haveData)
```

# The data

`ChrAccRex` ships an ATAC-seq example dataset of sorted immune cells,
aligned to hg38 and stored as a `DsATAC` object with fragment level
data. That is what `chromTFR` needs: footprints are built from single
base Tn5 cut positions, which only fragments provide.

```{r install, eval = FALSE}
remotes::install_github("EpigenomeInformatics/ChrAccRex")
```

```{r load}
library(chromTFR)
library(data.table)
library(ggplot2)

dsa <- ChrAccRex::loadExample("dsAtac_ia_example")
ann <- getAccSampleAnnotation(dsa)
table(ann$cellType)
ChrAccR::getRegionTypes(dsa)
```

This vignette compares naive against effector memory CD8 T cells, using
three unstimulated samples of each. Restricting to a handful of samples
keeps the build short; a real run would use all of them.

```{r groups}
groups <- c("TCD8naive", "TCD8EM")
picked <- do.call(rbind, lapply(groups, function(g) {
    rows <- ann[ann$cellType == g & ann$stimulus == "U", , drop = FALSE]
    head(rows, 3)
}))
picked$sampleId
```

# Insertion sites

The peaks of the example dataset are extended so that they still cover
the flanks of a motif window, and the insertions are counted per base
inside them. Scores are insertions per million, so samples of different
depth stay comparable.

```{r insertions}
peaks <- getAccRegions(dsa, regionType = "IA_prog_peaks", extend = 500)

cache <- tempfile("chromTFR_example")
dir.create(cache)
files <- vapply(picked$sampleId, function(s) {
    f <- file.path(cache, paste0("ins_", s, ".RDS"))
    saveRDS(getTn5Insertions(dsa, s, regions = peaks), f)
    f
}, character(1))
```

`mergeInsertionSites()` pools the samples of a group into one aggregate,
averaging the scores so that groups of unequal size stay on the same
scale.

```{r aggregate}
ins <- lapply(groups, function(g) {
    mergeInsertionSites(files[picked$cellType == g])
})
names(ins) <- groups
ins[["TCD8naive"]]
```

# Footprints

The annotation comes from `methylTFRAnnotationHg38`, unchanged from
`methylTFR`. `prepareTFBS()` resizes the binding sites to the footprint
window; everything downstream works on that object.

```{r annotation}
motifs <- c("CTCF", "FOS::JUN")
tf_bindsites <- methylTFRAnnotationHg38::getTFbindsites(
    motifSet = "jaspar2020"
)
gcfreqs <- methylTFRAnnotationHg38::getGCfreq(motifSet = "jaspar2020")
gc_dist <- methylTFRAnnotationHg38::getGenomeGC()

tfbs <- lapply(motifs, function(m) prepareTFBS(tf_bindsites[[m]]))
names(tfbs) <- motifs
lengths(tfbs)
```

The expected profile is the GC model here: the insertion density of the
sample summarised per GC bin, combined with the GC composition of the
binding sites.

```{r expected}
binIns <- lapply(ins, addGCBintoAccessome, gcdist = gc_dist)
expected <- lapply(groups, function(g) {
    prof <- lapply(motifs, function(m) {
        computeAccExpectations(binIns[[g]], gcfreqs[[m]])
    })
    names(prof) <- motifs
    prof
})
names(expected) <- groups
```

`accFootprintData()` stacks the insertions on the centre of every
binding site and returns the observed and expected curves of both
groups.

```{r profiles}
profiles <- lapply(motifs, function(m) {
    e <- lapply(groups, function(g) expected[[g]][[m]])
    names(e) <- groups
    accFootprintData(tfbs[[m]], ins, e)
})
names(profiles) <- motifs
head(profiles[["CTCF"]])
```

The deviation score reads that curve as one number, the insertion
density of the central window over the density of the outer flanks,
minus the same ratio on the expected curve. A value below zero means the
centre is more protected than the sequence composition predicts.

```{r deviations}
vapply(profiles, function(df) {
    df[group == "TCD8naive", accDeviation(.SD),
        .SDcols = c("x", "avg_ins", "type")]
}, numeric(1))
```

```{r panels, fig.width = 10, fig.height = 8}
colours <- c(TCD8naive = "#3B6EA5", TCD8EM = "#B23A48")
plotAccFootprintGrid(profiles, tfbs, colours, corrected = "GC corrected")
```

CTCF is the control to look at first: it binds stably, so its footprint
is deep and the corrected curve dips well below one. AP-1 is the
opposite case, a factor with a short residence time whose footprint
stays shallow even where it is active.

# The k-mer model

Tn5 prefers some short sequences over others, and a motif is a fixed
string of letters in the centre of every window, so part of any dip is
chemistry rather than protein. `computeKmerBias()` measures that
preference from the cuts themselves and `kmerBiasProfile()` turns it
into an expected curve. It needs the genome sequence.

```{r kmer, eval = FALSE}
library(BSgenome.Hsapiens.UCSC.hg38)
genome <- BSgenome.Hsapiens.UCSC.hg38

bg <- kmerBackground(peaks, genome, k = 6L)
bias <- lapply(ins, computeKmerBias,
    regions = peaks, genome = genome, k = 6L, bg = bg
)
expectedK <- lapply(groups, function(g) {
    prof <- lapply(motifs, function(m) {
        kmerBiasProfile(tfbs[[m]], genome, bias[[g]], k = 6L)
    })
    names(prof) <- motifs
    prof
})
names(expectedK) <- groups
```

Comparing the two models on the same motif is the honest check: a dip
that survives the k-mer correction is a footprint, one that flattens was
Tn5 preferring the flanking sequence.

```{r cleanup, include = FALSE}
unlink(cache, recursive = TRUE)
```

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
