---
title: "A worked example with the ChrAccR example data"
author: "Irem B. Gunduz"
package: chromTFR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{A worked example with the ChrAccR example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
haveData <- all(vapply(
    c("ChrAccR", "ChrAccRex", "methylTFRAnnotationHg38",
        "BSgenome.Hsapiens.UCSC.hg38"),
    requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(eval = haveData)
```

# The data

`ChrAccRex` ships an ATAC-seq example dataset of sorted immune cells,
aligned to hg38 and stored as a `DsATAC` object with fragment level
data. That is what `chromTFR` needs: footprints are built from single
base Tn5 cut positions, which only fragments provide.

```{r install, eval = FALSE}
remotes::install_github("EpigenomeInformatics/ChrAccRex")
```

```{r load}
library(chromTFR)
library(data.table)
library(ggplot2)
library(BSgenome.Hsapiens.UCSC.hg38)

dsa <- ChrAccRex::loadExample("dsAtac_ia_example")
ann <- getAccSampleAnnotation(dsa)
# The annotation is in the order of the samples, but its sampleId column
# lists the merged replicates, so take the ids from the object itself
ann$sample <- getAccSamples(dsa)
table(ann$cellType)
ChrAccR::getRegionTypes(dsa)
```

This vignette compares naive against effector memory CD8 T cells, using
three unstimulated samples of each. Restricting to a handful of samples
keeps the build short; a real run would use all of them.

```{r groups}
groups <- c("TCD8naive", "TCD8EM")
picked <- do.call(rbind, lapply(groups, function(g) {
    rows <- ann[ann$cellType == g & ann$stimulus == "U", , drop = FALSE]
    head(rows, 3)
}))
if (!all(groups %in% picked$cellType)) {
    stop("No unstimulated samples for: ",
        paste(setdiff(groups, picked$cellType), collapse = ", ")
    )
}
picked$sample
```

# Insertion sites

The peaks of the example dataset are extended so that they still cover
the flanks of a motif window, and the insertions are counted per base
inside them. Scores are insertions per million, so samples of different
depth stay comparable.

```{r insertions}
peaks <- getAccRegions(dsa, regionType = "IA_prog_peaks", extend = 500)

cache <- tempfile("chromTFR_example")
dir.create(cache)
files <- vapply(picked$sample, function(s) {
    f <- file.path(cache, paste0("ins_", s, ".RDS"))
    saveRDS(getTn5Insertions(dsa, s, regions = peaks), f)
    f
}, character(1))
```

`mergeInsertionSites()` pools the samples of a group into one aggregate,
averaging the scores so that groups of unequal size stay on the same
scale.

```{r aggregate}
ins <- lapply(groups, function(g) {
    mergeInsertionSites(files[picked$cellType == g])
})
names(ins) <- groups
ins[["TCD8naive"]]
```

# Binding sites

The annotation comes from `methylTFRAnnotationHg38`, unchanged from
`methylTFR`. `prepareTFBS()` resizes the binding sites to the footprint
window; everything downstream works on that object.

```{r annotation}
motifs <- c("CTCF", "FOS::JUN")
tf_bindsites <- methylTFRAnnotationHg38::getTFbindsites(
    motifSet = "jaspar2020"
)

tfbs <- lapply(motifs, function(m) prepareTFBS(tf_bindsites[[m]]))
names(tfbs) <- motifs
lengths(tfbs)
```

# The Tn5 k-mer model

Tn5 prefers some short sequences over others, and a motif is a fixed
string of letters in the centre of every window, so part of any dip is
chemistry rather than protein. The expected profile here measures that
preference from the cuts themselves.

`kmerBackground()` counts every 6-mer of the accessible genome. It
depends on the regions only, so it is computed once and reused for both
groups.

```{r background}
genome <- BSgenome.Hsapiens.UCSC.hg38
kmer <- 6L
bg <- kmerBackground(peaks, genome, k = kmer)
```

`computeKmerBias()` compares the 6-mer at each observed cut with that
background. A weight above one means Tn5 cuts that word more often than
its abundance predicts.

```{r bias}
bias <- lapply(ins, computeKmerBias,
    regions = peaks, genome = genome, k = kmer, bg = bg, max.ins = 5e5
)
summary(bias[["TCD8naive"]])
```

`kmerBiasProfile()` walks along the binding sites, reads the genomic
6-mer at each position and averages its weight across sites. That curve
is what the insertions would look like with no protein bound.

```{r expected}
expected <- lapply(groups, function(g) {
    prof <- lapply(motifs, function(m) {
        kmerBiasProfile(tfbs[[m]], genome, bias[[g]],
            k = kmer, max.sites = 5000
        )
    })
    names(prof) <- motifs
    prof
})
names(expected) <- groups
```

# Footprints

`accFootprintData()` stacks the insertions on the centre of every
binding site and matches the expected curve to the observed flanks.

```{r profiles}
profiles <- lapply(motifs, function(m) {
    e <- lapply(groups, function(g) expected[[g]][[m]])
    names(e) <- groups
    accFootprintData(tfbs[[m]], ins, e)
})
names(profiles) <- motifs
head(profiles[["CTCF"]])
```

The deviation score reads that curve as one number, the insertion
density of the central window over the density of the outer flanks,
minus the same ratio on the expected curve. A value below zero means the
centre is more protected than the sequence alone predicts.

```{r deviations}
vapply(profiles, function(df) {
    df[group == "TCD8naive", accDeviation(.SD),
        .SDcols = c("x", "avg_ins", "type")]
}, numeric(1))
```

```{r panels, fig.width = 10, fig.height = 8}
colours <- c(TCD8naive = "#3B6EA5", TCD8EM = "#B23A48")
plotAccFootprintGrid(profiles, tfbs, colours,
    corrected = "k-mer corrected"
)
```

CTCF is the control to look at first: it binds stably, so its footprint
is deep and the corrected curve dips well below one. AP-1 is the
opposite case, a factor with a short residence time whose footprint
stays shallow even where it is active.

A GC based expected profile is available as well, through
`addGCBintoAccessome()` and `computeAccExpectations()`, and needs no
genome sequence. Comparing the two on the same motif is a useful check:
a dip that survives the k-mer correction is a footprint, one that
flattens was Tn5 preferring the flanking sequence.

```{r cleanup, include = FALSE}
unlink(cache, recursive = TRUE)
```

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
