---
title: "Transcription factor footprints in chromatin accessibility data"
author: "Irem B. Gunduz"
package: chromTFR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{chromTFR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`chromTFR` is the chromatin accessibility counterpart of `methylTFR`. It
uses the same annotation, but methylation levels are replaced by Tn5
insertion densities: for every binding site of a motif the insertions are
counted per base, the sites are stacked on their centre and the average
gives the footprint. A protein bound to the motif shields the DNA from
Tn5, so the profile dips in the centre and rises on the flanks.

The deviation score is that shape as one number, the insertion density of
the central window over the density of the outer flanks, corrected by the
same ratio computed on an expected profile.

# The expected profile

Two models are available.

The GC model follows `methylTFR`: the genome wide insertion density is
summarised per GC bin with `addGCBintoAccessome()` and combined with the
GC bin composition of the binding sites by `computeAccExpectations()`.

The k-mer model captures the sequence preference of Tn5 directly.
`computeKmerBias()` counts the k-mer centred on every observed cut and
compares it with the frequency of that k-mer in the accessible genome,
and `kmerBiasProfile()` averages those weights along the binding sites.
The motif is a fixed string of letters at the centre of every window, so
this term matters: for an AT-rich motif Tn5 cuts less in the centre for
chemical reasons alone, which looks like protection.

# A minimal workflow

```{r workflow, eval = FALSE}
library(chromTFR)

dsa <- loadAccDataset("dsATAC_filtered")
peaks <- getAccRegions(dsa, regionType = ".peaks.cons", extend = 500)
ins <- getTn5Insertions(dsa, getAccSamples(dsa)[1], regions = peaks)

tfbs <- prepareTFBS(tf_bindsites[["CTCF"]])
profile <- accProfile(ins, tfbs)
accDeviationScore(profile)
```

`runChromTFR()` runs the GC based workflow over every sample and motif
and returns a `SummarizedExperiment` with the deviations, their row-wise
z-scores and the expected deviations.

# Figures

`accFootprintData()` builds the observed and expected curves of one motif
for a named list of samples or group aggregates, and
`plotAccFootprintGrid()` draws them as two rows, the profiles on top and
the corrected curves below, each labelled with its deviation score.

For a per sample overview, `accObservedMatrix()` and
`accExpectedMatrix()` give the two halves of the deviation matrix and
`plotAccDeviationHeatmap()` draws the row-wise z-scores. Fit the Tn5
model per sample there: with one shared model the expected term is
constant per motif and cancels out of a row-wise z-score.

# Analysis scripts

Template scripts for the whole workflow are installed with the
package. Copy one, set the paths and the grouping column at the top,
and run it.

```{r scripts, eval = FALSE}
list.files(system.file("scripts", package = "chromTFR"))
```

# Session info

```{r session}
sessionInfo()
```
